---
title: "Getting and Cleaning Data Course Project"
output: 
  html_document:
    keep_md: true
---

##Upload data into R

```{r}
X_train<-read.table("X_train.txt")
y_train<-read.table("y_train.txt")
X_test<-read.table("X_test.txt")
y_test<-read.table("y_test.txt")
subject_test<-read.table("subject_test.txt")
subject_train<-read.table("subject_train.txt")
feature_labs<-read.table("features.txt")
activity_labels<-read.table("activity_labels.txt")
```


##1. Merge the test and training datasets

First, check that the feature labels match X_test & X_train column names.

```{r}
feature_labs$V3<-paste0("V",feature_labs$V1)
sum(!(names(X_train)==feature_labs$V3))
sum(!(names(X_test)==feature_labs$V3))
```

Next, change the column names of X_test & X_train to the feature names.

```{r}
names(X_train)<-feature_labs$V2
names(X_test)<-feature_labs$V2
```

Attach the activity codes and subject labels to X_test & X_train.

```{r}
X_train2<-cbind(activitycode=y_train$V1,X_train)
X_train2<-cbind(subject=subject_train$V1,X_train2)
X_test2<-cbind(activitycode=y_test$V1,X_test)
X_test2<-cbind(subject=subject_test$V1,X_test2)
```

Check that the colum names of training and testing datasets are equivalent.

```{r}
sum(!(names(X_train2)==names(X_test2)))
```

Finally, merge the training and testing datasets to form one data.frame.

```{r}
complete<-rbind(X_test2,X_train2)
```


##2. Extract the mean and standard deviation of measurements

Note: featuresinfo.txt explains that each measured signal had an estimated mean (mean()) and standard deviation (std()). In this step, only features that contained "mean()" or "std()" were extracted. I did not extract the meanFreq() or the angle() means as these appeared to be additional calculations different from the mean and standard deviation of the measured signals.

Extract a subset of the data that contains either measurement mean or standard deviation and add the "subject" and "activitycode" columns to new subsetted dataset.

```{r}
complete2<-complete[,grepl("mean\\(\\)|std\\(\\)",names(complete))]
complete2<-cbind(complete[,1:2],complete2)
```


##3. Name the activities with descriptive names

Clean up labels by removing "_"'s and changing to lowercase letters.

```{r}
activity_labels$V2<-gsub("_","",activity_labels$V2)
activity_labels$V2<-tolower(activity_labels$V2)
names(activity_labels)<-c("activitycode","activity")
```

Merge the activity labels data.frame with the complete dataset to add a column with activity names.

```{r}
complete3<-merge(activity_labels,complete2,all=TRUE)
```


##4. Label the dataset with descriptive variable names

Note: The descriptive variable names provided with the dataset (features.txt) were added to the dataset previously in step #1. Descriptions of these variable names can be found in project_codebook.txt.

Cleanup column names by removing dashes and parentheses. 

```{r}
new_names<-gsub("-","_",names(complete3))
new_names<-gsub("\\(|\\)","",new_names)
names(complete3)<-new_names
```


##5. Create a second tidy data set with the average of each variable for each activity and each subject

Check for the presence of NA's.

```{r}
sum(is.na(complete3))
```

Create a new data.frame with the subject by activity averages using the aggregate function and write the new data.frame to a text file.

```{r}
project.data<-aggregate(complete3[,-c(1:3)],list(subject=complete3$subject,activity=complete3$activity),FUN=mean)
write.table(project.data,file="project.data.txt",row.name=FALSE)
```
